Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-10-16 manuscript/climate_sensitivity_draft.Rmd
Added all references from introduction, and updated text with citations

\newpage
### Introduction
Forests globally play a critical role in climate regulation ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)), yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change ([Friedlingstein et al. 2006](https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1)). An important aspect of this uncertainty lies in responses to drought (**REF**). In many forested regions around the world, the risk of severe drought is increasing ([Trenbert et al. 2014](https://www.nature.com/articles/nclimate2067)), even in conjunction with increasing precipitation ([IPCC 2014](https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334)). Global change-type drought has been affecting forests worldwide ([Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001)), and it is expected that future climate change-driven droughts will severely impact forests around the world ([Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**). Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [Stovall et al. 2019]( https://doi.org/10.1038/s41467-019-12380-6)), resulting in disproportionate impacts on forest C storage ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear. Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al. 2019, Hacket-Pain et al. 2016 (DOI 10.1007/s10342-016-0982-7) **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E. M., Hicke, J. A. & Meddens, A. J. H. Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States. Glob. Change Biol. 17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N. G., Bond, B. J., Hill, L., Ryan, M. G. & Whitehead, D. in Size and age related changes in tree structure and function (eds Meinzer, F.C. & Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N. G. & Allen, C. D. Darcy’s law predicts widespread forest mortality under climate warming. Nat. Clim. Change 5, 669–672 (2015); Ryan, M. G., Phillips, N. & Bond, B. J. The hydraulic limitation hypothesis revisited. Plant Cell Environ. 29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al. 2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)). Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al. 2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand ([D'Orangeville et al. 2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096)). However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. (**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011];  D'Orangeville et al. 2018 **REFS**), these don't test for differential trait effects across droughts (D'Orangeville et al. 2018) and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; Stovall et al. 2019). [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
(**PELTIER**)

@@

\newpage
### Supplementary Information
```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
tableS2 <- read.csv("tables_figures/candidate_traits.csv", stringsAsFactors = FALSE)
kable(tableS2, booktabs=TRUE, format = "latex", caption = "Candidate variables for best model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

```{r, Plot, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
```

*how do we want to present Table S3? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/top_models_dAIC.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Top model variations for each drought scenario, with dAICc values <= 2") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

![Location of cored trees](tables_figures/FigureS3.png){width=500px}

![Figure S1](tables_figures/FigureS1.png)
(see Issue #32)

>>>>

\newpage
### Introduction
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies in responses to drought (**REF**). In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**). Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [@stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear. Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019] *(Elliot et al. 2015, Friedrichs et al. 2009) unsure which ones these are* have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019]. There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought [@anderegg_hydraulic_2018] **review this pub**, but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. (**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; Stovall; **MORE REFS**). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
(**PELTIER**)

@@
